Mapping Mixtures of Genotype-by-Weather Interactions in Switchgrass (Panicum virgatum)

### 2024-02-21 manuscript/revision2/revision2.qmd
manuscript: M&M edits and some QTL incorporation

We next looked for evidence of GxWeather by jointly re-estimating genetic effects across all eight common gardens using the mash model, which allows effects to covary between gardens in many ways (@fig-map D, SI Appendix, Datasets 1-6). To do this, we used GWAS on site-specific BLUPs for the timing of vegetative growth and flowering to determine preliminary single nucleotide polymorphism (SNP) effect estimates for both traits at each garden. We then jointly modeled genetic effect estimates, and the mixture of ways these effects might covary, across all eight common gardens, using a two-part procedure. We first modeled genetic effects with different combinations of covariance structures, using a subset of randomly selected SNPs used in the GWAS and their corresponding effect size estimates. In these models (hereafter referred to as "random"), we then specified a variety of potential covariance structures among genetic effect size estimates from each garden (as expanded on below) and used a greedy algorithm to iteratively select and add covariance structures that significantly improved the log likelihood of the *mash* model. Given the high correlation between some of the covariance patterns we tested, this approach allowed us to select only the subset of patterns of covariance that significantly improved the fit of the jointly inferred structure of genetic effect estimates present in the data. We then selected the set of SNPs and corresponding effect size estimates that had the largest effect estimates in univariate GWAS (hereafter "strong" effects) and jointly re-estimated these effect size estimates at all eight sites using the set of significant covariance matrices that most improved the model fit on the random effects.

We specified three categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1, SI Appendix, Section S2). For example, GxWeather covariance between gardens in MI and MO, for rainfall in the seven days prior to flowering, captures the covariance in the amount of rainfall genotypes A, B, and C receive in the seven days prior to their three flowering dates in MI and their three flowering dates in MO. If these six rainfall values are similar, covariance will be near one; if they are uncorrelated, covariance will be near zero; if there is sufficient rank changing in rainfall amounts between gardens for these genotypes, this covariance will be near negative one. We specified multiple covariance structures for the canonical, data-driven, and GxWeather categories, each of which represented a specific pattern of covariance in genetic effect size estimates that could be included in the model.

The GxWeather covariance structures allow hypothesis testing of specific weather variables as cues for the start of vegetative and reproductive growth. Say that a SNP in FLC controls flowering in a photoperiod-dependent manner. In that case, the joint estimate of effects for that SNP could have a high mixture proportion, or mass, on a covariance matrix created using a photoperiod-based environmental cue, such as day length two days prior to flowering. In our data, we would infer that the effect of that SNP on flowering was caused by a response to the environmental cue used to construct the GxWeather covariance structure with the largest mass. Overall, loadings of genetic effects on these GxWeather matrices in the model, along with the posterior effect size estimates provide information on genome-wide patterns of SNP-environment interaction.

The phenotypic correlations for the timing of vegetative growth had moderate negative correlations between the Texas and North gardens, particularly in the Gulf subpopulation and Both subpopulations (@fig-map B). If these phenotypic correlations have a genetic component, then they may be controlled by SNP effects that differed in sign across these regions (effects with antagonistic pleiotropy), SNP effects that are large in one garden or region and non-significant in others (effects with amplification), or a combination of these patterns. Five GxWeather covariance structures were selected by the greedy algorithm, one to two per subpopulation. Of these five, three had mass on them in mash models of the strong effects (@fig-covar A-B). Two of these three matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while one had all positive or near-zero covariances. SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the mash model of Gulf vegetative growth fell on a covariance matrix constructed using the covariance of daylength 14 days prior to the date of vegetative growth. The covariance matrix for this weather cue was very similar to the pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). Mash models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C).

@@

We next characterized the pairwise patterns of effects where we were confident in the sign of the effect at both gardens. We used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for antagonistic pleiotropy, or a sign change between conditions, carry an equal statistical burden to those for effects with the same sign. We also summarized the pairwise patterns of effect sign and magnitude within and between the Texas and North regions, within the natural ranges of the Gulf and Midwest subpopulations (\@fig-effects).

For greenup date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited antagonistic pleiotropy, or a difference in effect sign, between pairs of Texas and North gardens (\@fig-effects A). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (\@fig-effects). The majority of pairwise effects for greenup for the Midwest (\>55%) and Both (\>85%) subpopulations were the same sign, and effects frequently differed in magnitude between the MO and OK garden and other gardens (\@fig-effects X).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited antagonistic pleiotropy, or a difference in effect sign, within or between regions (\@fig-effects). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; \@fig-effects X). The Midwest population had relatively few significant effects for flowering, but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (\@fig-effects). Most differences in sign were between TX1, the southernmost garden, and all other gardens (\@fig-effects B). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

We sought additional experimental support for our joint re-estimates of SNP effects using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (\@fig-qtl A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of four environmental cues that we also used as covariance matrices in mash, and identified eight QTL for flowering date, six QTL for flowering GDD, ten QTL for flowering day length, and eight QTL flowering day length change, all of which showed QTL by environment interactions (SI Appendix, Fig. S3). All QTL for flowering overlapped one or more homologs from rice or *A. thaliana* with functionally validated roles in flowering (SI Appendix, Dataset 7). All flowering QTL intervals contained at least one SNP significant in at least one mash run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of mash SNPs in the 1% tail of significance (the 'mash 1% tail') within each QTL interval. At the 5% level, five QTL had enrichments of SNPs in the mash 1% tail. Overall, there were eleven significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the mash 1% tail in the QTL intervals. Our QTL intervals had more enrichments of SNPs in the mash 1% tail than were found for all but eleven of these sets of random genomic intervals (\@fig-qtl C, p = 0.011). Thus, we were able to experimentally support our joint re-estimates of SNP effects with a QTL mapping experiment using a separate mapping population.

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both subpopulations revealed substantial antagonistic pleiotropy in effects between the Texas and North gardens (\@fig-effects A). This result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic], and is the first experimental work using QTL mapping and GWAS across common gardens to find antagonistic pleiotropy to be common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

Our analysis of the timing of flowering showed that the Gulf and Midwest subpopulations have distinct GxWeather: flowering timing in the Midwest subpopulation has photoperiod-related variation, in that flowering timing covaries with a day length change signal two days before flowering occurs. In contrast, the Gulf subpopulation does not have variation in flowering based on a photoperiod cue. Instead, the Gulf subpopulation has variation in flowering that covaries with the rainfall that occurs in the week prior to flowering. The genomic regions affecting flowering that we re-estimated across all eight sites were also supported by QTL from an independent mapping population at these sites (\@fig-qtl C).

Identifying the environmental cues that are predictive of, or even correlated with, plant phenotypic responses remains a major challenge to studies interrogating gene action across many natural environments. The GxWeather photoperiod and cumulative rainfall cues we identify here are functions of the genotypes measured and capture only a minority of SNP effects on flowering. We know still less about the overwintering parameters that cause variation in the start of vegetative. Though we could assign SNP effects to a GxWeather covariance structure in the Gulf subpopulation, we could not do this for the Midwest subpopulation. More generally, it is difficult to predict the time scales over which individuals may integrate environmental cues, particularly in perennial species which may integrate these cues over longer time scales. If this integration time itself varies between individuals, we cannot select a covariance structure that reflects this, though these structures would likely be highly correlated with GxWeather structures we did include. Mash offers an opportunity to specify multiple environmental cues and compete them to explain patterns of genetic effects, allowing us to detect how important these cues are genome-wide, and how strongly each cue influences each SNP. This is a key development to further improve our understanding of genetic variation in GxE.

@@

![Comparisons of pairs of jointly re-estimated effects across eight gardens, for effects with lfsr \< 0.05. AHM is considering moving this figure to the supplement and substantially simplifying the figure in the main text, perhaps just showing total effects in Texas, the North, and maybe Oklahoma. Essentially making Table 2 into a new, simplified figure. Effect patterns for re-estimated SNP effects that are significant at pairs of sites by the local false sign rate test (lfsr), for (a) greenup date and (b) flowering date, within and between two genetic subpopulations. Common gardens are arranged in latitudinal order along the x- and y-axis. Effects in the first and fourth rows differ in sign at these pairs of gardens (p \< 0.05, lfsr). Effects in the second, third, fifth and sixth rows are identical in sign (p \< 0.05, lfsr). Effects in the second and fifth rows additionally differ in magnitude by a factor of \>0.4, while effects in the third and sixth row were not distinguishable by magnitude nor sign of the effect.](images/Figure_3_Effect_Types_by_Site.png){#fig-effects}

@@

To evaluate the prevalence and kinds of covariance patterns of SNP effects across our common gardens, we used multivariate adaptive shrinkage (*mash*) on SNP effect estimates from the diversity panel (35). Mash is a statistical method that allows estimation and comparison of many effects jointly across many different conditions; it improves on previous methods by allowing multiple, arbitrary correlations in effect sizes among conditions (SI Appendix, Section S5). To obtain SNP effect estimates, we first conducted univariate genome-wide association at each common garden for green-up and flowering date. We then analyzed SNP effects for the top 19K relatively unlinked (r^2^ \< 0.2) SNPs per condition using mash, as in @lovell2021genomic. Details on these models can be found in (SI Appendix, Section S6).

>>>>

We next looked for evidence of GxWeather by jointly re-estimating genetic effects across all eight common gardens using the mash model, which allows effects to covary between gardens in many ways (@fig-map D, SI Appendix, Datasets 1-6). We specified three categories of covariance structures across gardens: "canonical" covariance, with simple patterns of effect size covariance introduced in the initial *mash* manuscript; "data-driven" covariances derived from common patterns of SNP effects observed in the data, and "GxWeather" covariance, estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1, SI Appendix, Section S2). For example, the more similar that the amount of rainfall prior to flowering is for genotypes A and B between gardens in MI and MO, the nearer these covariances will be to one. If rainfalls are uncorrelated, covariance will be near zero; if there is rank changing (e.g. high rainfall before flowering of A at MO but not MI, and high rainfall before flowering for B at MI but not MO), this covariance will be near negative one. We specified multiple covariance structures for the canonical, data-driven, and GxWeather categories, each of which represented a specific pattern of covariance in genetic effect size estimates that could be included in the model.

The GxWeather covariance structures allow hypothesis testing of specific weather variables as cues for the start of vegetative and reproductive growth. Say that a SNP in FLC controls flowering in a photoperiod-dependent manner. In that case, the joint estimate of effects for that SNP could have a high mixture proportion, or mass, on a covariance matrix created using a photoperiod-based environmental cue, such as day length at some interval prior to flowering. In our data, we would infer that the effect of that SNP on flowering was caused by a response to the environmental cue used to construct the GxWeather covariance structure with the largest mass. Overall, loadings of genetic effects on these GxWeather matrices in the model, along with the posterior effect size estimates, provide information on genome-wide patterns of SNP-environment interaction.

The phenotypic correlations for the timing of vegetative growth had moderate negative correlations between the Texas and North gardens, particularly in the Gulf subpopulation and Both subpopulations (@fig-map B). If these phenotypic correlations have a genetic component, then they may be controlled by SNP effects that differ in sign across these regions (effects with antagonistic pleiotropy), SNP effects that are large in one garden or region and non-significant in others (effects with amplification), or a combination of these patterns. Five GxWeather covariance structures were selected by the greedy algorithm, one to two per subpopulation. Of these five, three had mass on them in mash models of the strong effects (@fig-covar A-B). Two of these three matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while one had all positive or near-zero covariances. SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the mash model of Gulf vegetative growth fell on a covariance matrix constructed using the covariance of daylength 14 days prior to the date of vegetative growth. The covariance matrix for this weather cue was very similar to the pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). Mash models of the timing of vegetative growth in the Midwest subpopulation and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C).

@@

We next characterized the pairwise patterns of effects where we were confident in the sign of the effect at both gardens. We used the local false sign rate (lfsr), an analogue of the lfdr that establishes confidence in the effect sign, not the effect's difference from zero, to determine significance. We required lfsr significance (p \< 0.05) in both gardens to include effects. This means that our tests for antagonistic pleiotropy, or a sign change between conditions, carry an equal statistical burden to those for effects with the same sign. We also summarized the pairwise patterns of effect sign and magnitude within and between the Texas and North regions, within the natural ranges of the Gulf and Midwest subpopulations (@fig-effects).

For greenup date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited antagonistic pleiotropy, or a difference in effect sign, between pairs of Texas and North gardens (@fig-effects A). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (@fig-effects). The majority of pairwise effects for greenup for the Midwest (\>55%) and Both (\>85%) subpopulations were the same sign, and effects frequently differed in magnitude between the MO and OK garden and other gardens (@fig-effects X).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited antagonistic pleiotropy, or a difference in effect sign, within or between regions (@fig-effects). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; @fig-effects X). The Midwest population had relatively few significant effects for flowering, but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (@fig-effects). Most differences in sign were between TX1, the southernmost garden, and all other gardens (@fig-effects B). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

We sought additional experimental support for our joint re-estimates of SNP effects using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (@fig-qtl A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of five environmental cues that we also used as covariance matrices in mash, and identified eight QTL for flowering date, eight QTL for flowering as a function of day length change two days prior, one QTL for the start of vegetative growth, and two QTL for vegetative growth as a function of daylength change one day prior, all of which showed QTL by environment interactions (SI Appendix, Fig. S3). All QTL for flowering overlapped one or more homologs from rice or *A. thaliana* with functionally validated roles in flowering (SI Appendix, Dataset 7). All flowering and green-up QTL intervals contained at least one SNP significant in at least one mash run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of mash SNPs in the 1% tail of significance (the 'mash 1% tail') within each QTL interval. At the 5% level, three QTL had enrichments of SNPs in the mash 1% tail. Overall, there were five significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the mash 1% tail in the QTL intervals; however, this number of enrichments was found in many random genomic intervals also (@fig-qtl C). Thus, we were able to experimentally support some of our re-estimates of SNP effects with a QTL mapping experiment using a separate mapping population.

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both subpopulations revealed substantial antagonistic pleiotropy in effects between the Texas and North gardens (\@fig-effects A). This result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic], and is the first experimental work using GWAS across common gardens to find antagonistic pleiotropy to be common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

Our analysis of the timing of flowering showed that the Gulf and Midwest subpopulations have distinct GxWeather: flowering timing in the Midwest subpopulation has photoperiod-related variation, in that flowering timing covaries with a day length change signal two days before flowering occurs. In contrast, the Gulf subpopulation does not have variation in flowering based on a photoperiod cue. Instead, the Gulf subpopulation has variation in flowering that covaries with the rainfall that occurs in the week prior to flowering. Three genomic regions affecting flowering that we re-estimated across all eight sites were also supported by QTL from an independent mapping population at these sites (\@fig-qtl A).

Identifying the environmental cues that are predictive of, or even correlated with, plant phenotypic responses remains a major challenge to studies interrogating gene action across many natural environments. The GxWeather photoperiod and cumulative rainfall cues we identify here are functions of the genotypes measured and capture only a minority of SNP effects on flowering. We know still less about the overwintering parameters that cause variation in the start of vegetative growth. Though we could assign SNP effects to a GxWeather covariance structure in the Gulf subpopulation, we could not do this for the Midwest subpopulation. More generally, it is difficult to predict the time scales over which individuals may integrate environmental cues, particularly in perennial species which may integrate these cues over longer time scales. If this integration time itself varies between individuals, we cannot select a covariance structure that reflects this, though these structures would likely be highly correlated with GxWeather structures we did include. Mash offers an opportunity to specify multiple environmental cues and compete them to explain patterns of genetic effects, allowing us to detect how important these cues are genome-wide, and how strongly each cue influences each SNP. This is a key development to further improve our understanding of genetic variation in GxE.

@@

![ AHM is considering moving this figure to the supplement and substantially simplifying the figure in the main text, perhaps just showing total effects in Texas, the North, and maybe Oklahoma. Comparisons of pairs of jointly re-estimated effects across eight gardens, for effects with lfsr \< 0.05. Essentially making Table 2 into a new, simplified figure. Effect patterns for re-estimated SNP effects that are significant at pairs of sites by the local false sign rate test (lfsr), for (a) greenup date and (b) flowering date, within and between two genetic subpopulations. Common gardens are arranged in latitudinal order along the x- and y-axis. Effects in the first and fourth rows differ in sign at these pairs of gardens (p \< 0.05, lfsr). Effects in the second, third, fifth and sixth rows are identical in sign (p \< 0.05, lfsr). Effects in the second and fifth rows additionally differ in magnitude by a factor of \>0.4, while effects in the third and sixth row were not distinguishable by magnitude nor sign of the effect.](images/Figure_3_Effect_Types_by_Site.png){#fig-effects}

@@

To evaluate the prevalence and kinds of covariance patterns of SNP effects across our common gardens, we used multivariate adaptive shrinkage (*mash*) on SNP effect estimates from the diversity panel (35). To do this, we first conducted univarate genome-wide association (GWAS) on site-specific best linear unbiased predictors (BLUPs) for the timing of vegetative growth and flowering, using the switchgrassGWAS package and the methods in [@lovell2021]. We used the effect estimates for single nucleotide polymorphisms (SNPs) from these GWAS as preliminary effect estimates for both traits at each garden. We then jointly modeled genetic effect estimates, and the mixture of ways these effects might covary, across all eight common gardens, using a two-part procedure. We first modeled genetic effects with different combinations of covariance structures, using the effect size estimates from a subset of [19K]{.underline} relatively unlinked (r^2^ \< 0.2), randomly selected SNPs used in the GWAS. In these models (hereafter referred to as "random"), we then specified a variety of potential covariance structures among genetic effect size estimates from each garden (as expanded on below) and used a greedy algorithm to iteratively select and add covariance structures that significantly improved the log likelihood of the *mash* model. Given the high correlation between some of the covariance patterns we tested, this approach allowed us to select only the subset of patterns of covariance that significantly improved the fit of the jointly inferred structure of genetic effect estimates present in the data. We then selected 19K relatively unlinked (r^2^ \< 0.2) SNPs from each univariate GWAS that had the largest effect estimates in each GWAS (hereafter "strong" effects), which gave a set of [**48K - 80K**]{.underline} SNPs, and jointly re-estimated these effect size estimates at all eight sites using the set of significant covariance matrices that most improved the model fit on the random effects. Notably, the majority of these "strong" effects were not significant in univariate GWAS.
